### 2022-12-05 dissertation/chapters/03-heuristics.tex
💚 Dedicated to Taylah, Beané and Aleah
✨ (review): Spelling and grammar, derived publications fix

Optimisation is the task of finding a solution to a given problem that is better than alternative solutions. Better stated by Oldewage~\cite{ref:oldewage:2017}, optimisation is the task of finding values for a set of variables such that some measure of optimality is satisfied given a set of constraints. Engelbrecht~\cite{ref:engelbrecht:2007} breaks optimisations problems down into three components:

@@

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \acs{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sough out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic.

@@

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \acp{MH} make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \acp{MH}.

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.75\textwidth]{images/gradient_descent.pdf}
      \caption{An illustration of \acf{GD} over various timesteps showing the minimisation of the error with regards to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

@@

\begin{itemize}
      \item \textbf{Feedforward Pass}: During this phase the output values of the \acs{FFNN} is calculcated for each training pattern.

@@

where $w_{i}$ is the weight value at index $i$, $t$ is the timestep, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and timestep $t$. The delta weight term $\Delta w_{i}(t)$, is given as

@@

where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given as

@@

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the generic \acf{GD} heuristic.}
      \label{algo:heuristics:gd:gd}
      \begin{algorithmic}
            \State Initialise the \acs{FFNN} weights and biases, the learning rate $\eta$ and the timestep/epoch $T=0$;
            \While{stopping conditions are not met}
            \State Let $\epsilon_{T} = 0$;
            \For{each training pattern p}
            \State Calculate the output of the hidden and output layers (feedforward);
            \State Compute the error signals of the hidden and output layers;
            \State Adjust the weights of the output and hidden layers (backpropagate);
            \State $\epsilon_{T}$ += $[\epsilon_{T_{p}} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
            \EndFor
            \State $T += 1$;
            \EndWhile
      \end{algorithmic}
\end{algorithm}

@@

where $\boldsymbol{w}$ refers to the weight vector, $\eta$ refers to the learning rate as before and $\boldsymbol{g}$ refers to the gradient of the error function relative to weight vector. Note that some subscripts related to layers, indices, timesteps and training patterns are omitted for convenience.

@@

\acs{SGD} was one of the first widely used heuristics to train \acs{FFNN}, however, it is not without flaws. With stochastic training, only one training pattern is presented at each iteration/epoch. As such, weight updates are done with high variance and noise~\cite{ref:ruder:2016}. An illustration of the the fluctuations caused by \acs{SGD} during training is given in Figure~\ref{fig:heuristics:gd:sgd}.

@@

      \item It is difficult to avoid getting trapped in local minima, especially for highly non-convex loss functions used for training \acp{FFNN}. \citeauthor{ref:dauphin:2014}~\cite{ref:dauphin:2014} mentions the role of sadle points in getting trapped in local minima. Sadle points are points at which one dimensions slopes upwards, while another dimension slopes downwards. \citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that these sadle points are usually surrounded by plateaus of the same error, leading to gradients that are close to zero in all dimensions.
\end{itemize}

@@

\Acs{Momentum} ~\cite{ref:qian:1999} is a variant of \acs{SGD} that helps accelerate \acs{SGD} in the relevant direction, dampening oscillations.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$, of the weight update vector of the previous timestep to the current weight update vector. An illustration of \acs{SGD} with and without momentum is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

@@

\Acf{Adagrad} is an adaptation of \acs{SGD} that implements a learning rate for every element in the weight vector and is developed by Duchi et al.~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \acs{Adagrad} adapts the learning rate to the elements of the weight vector, by  performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason, \acs{Adagrad} is well suited for dealing with situations where training data is sparse.

In the \acs{GD} variants presented thus far, the weight updates have been applied by making use of the same learning rate $\eta$, for all elements of the weight update vector. \Acs{Adagrad} uses a different learning rate for every weight element $w_{i}$, at every timestep $t$. The weight update step for \acs{Adagrad} is given as

@@

where $w_{i}$ is the $i$-th element of the weight vector, $G_{t,ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$, is the sum of the squared gradients w.r.t. $w_{i}$, up to timestep $t$ and $\epsilon$ is a smoothing term that avoids division by zero and is set to an insignificant value such as $1 \times 10^{-8}$. Since $\boldsymbol{G_{t}}$ contains the sum of the squares of the gradients along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product, represented by $\odot$, between $G_{t,i}$ and $g_{t,i}$. The simplified update step as implemented by \acs{Adagrad} is given as

@@

Zeiler~\cite{ref:zeiler:2012} presents \acs{Adadelta}, an improvement of \acs{Adagrad} that eliminates the accumulation of squared gradients in the denominator.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \acs{Adagrad}, \acs{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average of squared gradients at timestep $t$, denoted by $E[\boldsymbol{g}^{2}]_{t}$  depends on a fraction $\alpha$ of the previous average of squared gradients and the current gradient, similar to the update step for \acs{Momentum}. Also similar to \acs{Momentum}, $\alpha$ is set to 0.9. The update step for $E[\boldsymbol{g}^{2}]_{t}$ is given as

@@

~\citeauthor{ref:zeiler:2012}~\cite{ref:zeiler:2012} noted that the units of the update step as presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} do not match. The weight update vector should have the same hypothetical units as the weight vector. To fix the aforemention problem, another exponentially decaying average is defined in terms of squared weight updates and is given as

@@

Note that $RMS[\Delta \boldsymbol{w}]_{t}$ is unknown at timestep $t$. $RMS[\Delta \boldsymbol{w}]_{t}$ is approximated with the \acf{RMS} of weight updates until the previous timestep. The learning rate in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} is then replaced. The weight update step for \acs{Adadelta} is then concluded and is given as

@@

\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that $\boldsymbol{m}_{t}$ and $\boldsymbol{v}_{t}$ presented above are estimates of the first moment (the mean) and the second moment (the uncentered variance) of the gradients respectively. ~\citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} mentions that because $\boldsymbol{m}_{t}$ and $\boldsymbol{v}_{t}$ are initialised to be vectors of 0's, they are biased towards 0. The aforementioned bias is especially prominent during the initial timesteps and/or when the decay rates $\beta_{1}$ and $\beta_{2}$ are small ($\beta_{1}$ and $\beta_{2}$ are close to 1). The bias-corrected first and second moment estimates are presented in Equations~\eqref{eq:heuristics:gd:adam_bc_first_moment} and~\eqref{eq:heuristics:gd:adam_bc_second_moment} respectively.

@@

The term \acf{MH} was first introduced by~\citeauthor{ref:glover:1986}~\cite{ref:glover:1986} in 1986 and is derived from the composition of two Greek words. \index{heuristic}Heuristic derives from the verb \textit{heuriskein} which means ``to find''. The prefix, \textit{meta}, means ``beyond, in an upper level''. \Acp{MH} were often called \textit{modern heuristics}~\cite{ref:reeves:1993}.~\citeauthor{ref:blum:2003}~\cite{ref:blum:2003} mention that there is a debate as to what the formal defintion of \acp{MH} is and suggests the definition of \acp{MH} to be ``high level strategies for exploring search spaces by using different methods''.

@@

This dissertation uses the term \textit{entity} for candidate solutions and a \textit{population} for a collection of entities, in the general context of population-based \acp{MH}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} mentions that in \acp{PSO} individual candidate solutions are referred to as \textit{particles} and the population is referred to as a \textit{swarm}. These particles are ``flown'' through a hyperdimensional search space. Changes in particle position is due to social-psychological tendencies of individuals to emulate the success of other individuals. The changes in the particle position are then influenced by the experience or knowledge of the particle's neighbours. The social behaviour of particles is modeled such that they stochastically return to previously successful regions in the search space.

@@

\begin{itemize}
      \item \textbf{Position}: Refers to the candidate solution that is represented by the particle and defines the particle position within the optimisation problem's hyper-dimensional solution space. Let the current position vector, for particle $i$, at timestep $t$ be denoted by $\boldsymbol{x}_{i}(t)$. Let $I$ denoted the population size and $J$ denote the search space dimensionality.

      \item \textbf{Velocity}: Represents a step size for the particle in the search space. The velocity at vector, for particle $i$, at timestep $t$ is denoted $\boldsymbol{v}_{i}(t)$.

@@

During initialisation, particles are randomly placed within the search space by sampling from a uniform distribution such that $\boldsymbol{x}_{i} \sim U(x_{min}, x_{max})$ and the velocity is set to 0. At timestep 0, the particle's initial position is set to be the particle's personal best solution such that $\boldsymbol{y}_{i}(0) = \boldsymbol{x}_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position}.

@@

~\citeauthor{ref:vanwyk:2014}\cite{ref:vanwyk:2014} mentions that when \acp{PSO} where first developed, it was possible for particle velocities to become inappropriately large during optimisation, leading to situations where particles fly out of the feasible search space. This is known as \textit{swarm explosion} and occurs when there are frequent changes in the global best position. In order to address this issues, the concept of \textit{velocity clamping} was introduced~\cite{ref:eberhart:1996}. The idea behind velocity clamping is to restrict particle velocities to some $V_{max}$ threshold, modeling a form of terminal velocity. Velocity clamping is applied after the velocity update step and is given in Equation~\eqref{eq:heuristics:pso:velocity_clamping} below.

\begin{equation}
      \label{eq:heuristics:pso:velocity_clamping}
      \begin{split}
            v_{ij}(t+1)=
            \begin{cases}
                  v'_{ij}(t+1) & \text{if } -V_{max,j} < v'_{ij}(t+1) < V_{max,j} \\
                  -V_{maxj}    & \text{if } v'_{ij}(t+1) \leq -V_{max,j}          \\
                  V_{maxj}     & \text{if } v'_{ij}(t+1) \geq V_{max,j}
            \end{cases}
      \end{split}
\end{equation}

@@

An example of the pseudo code implementation of the \textit{gbest} \acs{PSO} is taken from~\cite{ref:engelbrecht:2007} and is given in Algorithm~\ref{algo:heuristics:pso:gbest}.

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the gbest \acs{PSO} heuristic.}
      \label{algo:heuristics:pso:gbest}
      \begin{algorithmic}
            \State Create swarm of $N$ entities, each with $J$ dimensions;
            \While{stopping condition are not met}
            \For{each particle $i = 1, \dots, N$}
            \State // Set the personal best position;
            \If{$f(x_{i}) < f(y_{i})$ }
            \State $y_{i} = x_{i}$;
            \EndIf

@@

The purpose of the \index{mutation}mutation operator is to produce a trial vector for each entity in the current population by mutating a \textit{target} vector with a \textit{weighted differential}~\cite{ref:engelbrecht:2007}. The trial vector is used in the \index{crossover}crossover operator to produce offspring. The \index{mutation}mutation process then follows as such. For each each parent $\boldsymbol{x}_{i}(t)$ generate a trial vector $\boldsymbol{u}_{i}(t)$ as follows:

@@

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the binomial \index{crossover}crossover technique for \acs{DE}.}
      \label{algo:heuristics:de:bin}
      \begin{algorithmic}
            \State $j^{*} \sim U(1,N)$;
            \State $\mathcal{J} \gets \mathcal{J} \cup \{j^{*}\}$;
            \For{each $j \in \{1, \dots, N\}$}
            \If{$U(0,1) < p_{r}$ and $j \neq j^{*}$ }
            \State $\mathcal{J} \gets \mathcal{J} \cup \{j\}$;
            \EndIf
            \EndFor
      \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the exponential \index{crossover}crossover technique for \acs{DE}.}
      \label{algo:heuristics:de:exp}
      \begin{algorithmic}
            \State $\mathcal{J} \gets \{\}$;
            \State $j \sim U(0,N - 1)$;
            \Repeat
            \State $\mathcal{J} \gets \mathcal{J} \cup \{j + 1 \}$;
            \State $j = (j+1) \mod N$
            \Until{ $U(0,1) \geq p_{r}$ or $|\mathcal{J}| = N$;}
      \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[H]
      \caption{The pseudo code for the general \acs{DE} heuristic.}
      \label{algo:heuristics:de:general_de}
      \begin{algorithmic}
            \State Set the generation counter, $t = 0$;
            \State Initialise the control parameters, $\beta$ and $p_{r}$
            \While{stopping condition not met}
            \For{each entity $\boldsymbol{x}_{i}(t) \in \mathcal{C}(t)$ }
            \State Evaluate the fitness, $f(\boldsymbol{x}_{i}(t))$;
            \State Create the trial vector, $\boldsymbol{u}_{i}(t)$ by applying the mutation operator;
            \State Create an offspring, $\boldsymbol{x}'_{i}(t)$ by applying the crossover operator;
            \If{$f(\boldsymbol{x}'_{i}(t))$ is better than $f(\boldsymbol{x}_{i}(t))$ }
            \State Add $\boldsymbol{x}'_{i}(t)$ to $\mathcal{C}(t+1)$;
            \Else
            \State Add $\boldsymbol{x}_{i}(t)$ to $\mathcal{C}(t+1)$;
            \EndIf
            \EndFor
            \EndWhile
            \State Return the individual with the best fitness as the solution;
      \end{algorithmic}
\end{algorithm}

@@

Many variants of \acs{DE} have been created and researched~\cite{ref:mezura:2006}. A general notation for \acs{DE} heuristic variants have been developed by Storn et al.~\cite{ref:storn:1996, ref:storn:1997}. The notation follows the form \textit{DE/$x$/$y$/$z$} where $x$, $y$ and $z$ refer to the components that are used by the particular \acs{DE}. A breakdown of this notation is provided as follows:

@@

\Acp{GA} where first proposed by Fraser~\cite{ref:fraser:1957} and later by Bremermann~\cite{ref:bremermann:1962} and Reed et al.~\cite{ref:reed:1967}. However, Holland~\cite{ref:holland:1992} is widely regarded as the father of \acp{GA}. Similar to \acp{DE}, \acp{GA} are also nature-inspired population-based \acp{MH} and model genetic evolution of entities in a hypothetical population. As with other \acp{EA}, \acp{GA} implement a number of operators that drive the optimisation process.  Primarily, \acp{GA} implement selection, modeling survival of the fittest and \index{crossover}crossover, modeling reproduction. The remainder of this section presents the aforementioned operators in more detail.

>>>>

Optimisation is the task of finding a solution to a given problem that is better than alternative solutions. Better stated by \citeauthor{ref:oldewage:2017}~\cite{ref:oldewage:2017}, optimisation is the task of finding values for a set of variables such that some measure of optimality is satisfied given a set of constraints. \citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} breaks optimisations problems down into three components:

@@

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \acs{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sought out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic.

@@

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \acp{MH} make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This dissertation takes a particular interest in gradient-based \index{heuristic}heuristics and \acp{MH}.

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.75\textwidth]{images/gradient_descent.pdf}
      \caption{An illustration of \acf{GD} over various time steps showing the minimisation of the error with regards to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

@@

\begin{itemize}
      \item \textbf{Feedforward Pass}: During this phase the output values of the \acs{FFNN} is calculated for each training pattern.

@@

where $w_{i}$ is the weight value at index $i$, $t$ is the time step, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and time step $t$. The delta weight term $\Delta w_{i}(t)$, is given as

@@

where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each time step. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given as

@@

\begin{algorithm}[htb]
      \caption{The pseudo-code algorithm for the generic \acf{GD} heuristic.}
      \label{algo:heuristics:gd:gd}
      \begin{algorithmic}
            \State Initialise the \acs{FFNN} weights and biases, the learning rate $\eta$ and the time step/epoch $T=0$;
            \While{stopping conditions are not met}
            \State Let $\epsilon_{T} = 0$;
            \For{each training pattern p}
            \State Calculate the output of the hidden and output layers (feedforward);
            \State Compute the error signals of the hidden and output layers;
            \State Adjust the weights of the output and hidden layers (backpropagate);
            \State $\epsilon_{T}$ += $[\epsilon_{T_{p}} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
            \EndFor
            \State $T += 1$;
            \EndWhile
      \end{algorithmic}
\end{algorithm}

@@

where $\boldsymbol{w}$ refers to the weight vector, $\eta$ refers to the learning rate as before and $\boldsymbol{g}$ refers to the gradient of the error function relative to weight vector. Note that some subscripts related to layers, indices, time steps and training patterns are omitted for convenience.

@@

\acs{SGD} was one of the first widely used heuristics to train \acs{FFNN}, however, it is not without flaws. With stochastic training, only one training pattern is presented at each iteration/epoch. As such, weight updates are done with high variance and noise~\cite{ref:ruder:2016}. An illustration of the fluctuations caused by \acs{SGD} during training is given in Figure~\ref{fig:heuristics:gd:sgd}.

@@

      \item It is difficult to avoid getting trapped in local minima, especially for highly non-convex loss functions used for training \acp{FFNN}. \citeauthor{ref:dauphin:2014}~\cite{ref:dauphin:2014} mentions the role of saddle points in getting trapped in local minima. Saddle points are points at which one dimensions slopes upwards, while another dimension slopes downwards. \citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that these saddle points are usually surrounded by plateaus of the same error, leading to gradients that are close to zero in all dimensions.
\end{itemize}

@@

\Acs{Momentum} ~\cite{ref:qian:1999} is a variant of \acs{SGD} that helps accelerate \acs{SGD} in the relevant direction, dampening oscillations.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$, of the weight update vector of the previous time step to the current weight update vector. An illustration of \acs{SGD} with and without momentum is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

@@

\Acf{Adagrad} is an adaptation of \acs{SGD} that implements a learning rate for every element in the weight vector and is developed by \citeauthor{ref:duchi:2011}~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \acs{Adagrad} adapts the learning rate to the elements of the weight vector, by  performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason, \acs{Adagrad} is well suited for dealing with situations where training data is sparse.

In the \acs{GD} variants presented thus far, the weight updates have been applied by making use of the same learning rate $\eta$, for all elements of the weight update vector. \Acs{Adagrad} uses a different learning rate for every weight element $w_{i}$, at every time step $t$. The weight update step for \acs{Adagrad} is given as

@@

where $w_{i}$ is the $i$-th element of the weight vector, $G_{t,ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$, is the sum of the squared gradients w.r.t. $w_{i}$, up to time step $t$ and $\epsilon$ is a smoothing term that avoids division by zero and is set to an insignificant value such as $1 \times 10^{-8}$. Since $\boldsymbol{G_{t}}$ contains the sum of the squares of the gradients along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product, represented by $\odot$, between $G_{t,i}$ and $g_{t,i}$. The simplified update step as implemented by \acs{Adagrad} is given as

@@

\citeauthor{ref:zeiler:2012}~\cite{ref:zeiler:2012} presents \acs{Adadelta}, an improvement of \acs{Adagrad} that eliminates the accumulation of squared gradients in the denominator.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \acs{Adagrad}, \acs{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average of squared gradients at time step $t$, denoted by $E[\boldsymbol{g}^{2}]_{t}$  depends on a fraction $\alpha$ of the previous average of squared gradients and the current gradient, similar to the update step for \acs{Momentum}. Also similar to \acs{Momentum}, $\alpha$ is set to 0.9. The update step for $E[\boldsymbol{g}^{2}]_{t}$ is given as

@@

~\citeauthor{ref:zeiler:2012}~\cite{ref:zeiler:2012} noted that the units of the update step as presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} do not match. The weight update vector should have the same hypothetical units as the weight vector. To fix the aforementioned problem, another exponentially decaying average is defined in terms of squared weight updates and is given as

@@

Note that $RMS[\Delta \boldsymbol{w}]_{t}$ is unknown at time step $t$. $RMS[\Delta \boldsymbol{w}]_{t}$ is approximated with the \acf{RMS} of weight updates until the previous time step. The learning rate in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} is then replaced. The weight update step for \acs{Adadelta} is then concluded and is given as

@@

\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that $\boldsymbol{m}_{t}$ and $\boldsymbol{v}_{t}$ presented above are estimates of the first moment (the mean) and the second moment (the uncentered variance) of the gradients respectively. ~\citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} mentions that because $\boldsymbol{m}_{t}$ and $\boldsymbol{v}_{t}$ are initialised to be vectors of 0's, they are biased towards 0. The aforementioned bias is especially prominent during the initial time steps and/or when the decay rates $\beta_{1}$ and $\beta_{2}$ are small ($\beta_{1}$ and $\beta_{2}$ are close to 1). The bias-corrected first and second moment estimates are presented in Equations~\eqref{eq:heuristics:gd:adam_bc_first_moment} and~\eqref{eq:heuristics:gd:adam_bc_second_moment} respectively.

@@

The term \acf{MH} was first introduced by~\citeauthor{ref:glover:1986}~\cite{ref:glover:1986} in 1986 and is derived from the composition of two Greek words. \index{heuristic}Heuristic derives from the verb \textit{heuriskein} which means ``to find''. The prefix, \textit{meta}, means ``beyond, in an upper level''. \Acp{MH} were often called \textit{modern heuristics}~\cite{ref:reeves:1993}.~\citeauthor{ref:blum:2003}~\cite{ref:blum:2003} mention that there is a debate as to what the formal definition of \acp{MH} is and suggests the definition of \acp{MH} to be ``high level strategies for exploring search spaces by using different methods''.

@@

This dissertation uses the term \textit{entity} for candidate solutions and a \textit{population} for a collection of entities, in the general context of population-based \acp{MH}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} mentions that in \acp{PSO} individual candidate solutions are referred to as \textit{particles} and the population is referred to as a \textit{swarm}. These particles are ``flown'' through a hyper-dimensional search space. Changes in particle position is due to social-psychological tendencies of individuals to emulate the success of other individuals. The changes in the particle position are then influenced by the experience or knowledge of the particle's neighbours. The social behaviour of particles is modelled such that they stochastically return to previously successful regions in the search space.

@@

\begin{itemize}
      \item \textbf{Position}: Refers to the candidate solution that is represented by the particle and defines the particle position within the optimisation problem's hyper-dimensional solution space. Let the current position vector, for particle $i$, at time step $t$ be denoted by $\boldsymbol{x}_{i}(t)$. Let $I$ denoted the population size and $J$ denote the search space dimensionality.

      \item \textbf{Velocity}: Represents a step size for the particle in the search space. The velocity at vector, for particle $i$, at time step $t$ is denoted $\boldsymbol{v}_{i}(t)$.

@@

During initialisation, particles are randomly placed within the search space by sampling from a uniform distribution such that $\boldsymbol{x}_{i} \sim U(x_{min}, x_{max})$ and the velocity is set to 0. At time step 0, the particle's initial position is set to be the particle's personal best solution such that $\boldsymbol{y}_{i}(0) = \boldsymbol{x}_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position}.

@@

\citeauthor{ref:vanwyk:2014}~\cite{ref:vanwyk:2014} mentions that when \acp{PSO} where first developed, it was possible for particle velocities to become inappropriately large during optimisation, leading to situations where particles fly out of the feasible search space. This is known as \textit{swarm explosion} and occurs when there are frequent changes in the global best position. In order to address this issues, the concept of \textit{velocity clamping} was introduced~\cite{ref:eberhart:1996}. The idea behind velocity clamping is to restrict particle velocities to some $V_{max}$ threshold, modelling a form of terminal velocity. Velocity clamping is applied after the velocity update step and is given in Equation~\eqref{eq:heuristics:pso:velocity_clamping} below.

\begin{equation}
      \label{eq:heuristics:pso:velocity_clamping}
      \begin{split}
            v_{ij}(t+1)=
            \begin{cases}
                  v'_{ij}(t+1) & \text{if } -V_{max,j} < v'_{ij}(t+1) < V_{max,j} \\
                  -V_{max,j}   & \text{if } v'_{ij}(t+1) \leq -V_{max,j}          \\
                  V_{max,j}    & \text{if } v'_{ij}(t+1) \geq V_{max,j}
            \end{cases}
      \end{split}
\end{equation}

@@

An example of the pseudo-code implementation of the \textit{gbest} \acs{PSO} is taken from~\cite{ref:engelbrecht:2007} and is given in Algorithm~\ref{algo:heuristics:pso:gbest}.

\begin{algorithm}[htb]
      \caption{The pseudo-code algorithm for the gbest \acs{PSO} heuristic.}
      \label{algo:heuristics:pso:gbest}
      \begin{algorithmic}
            \State Create swarm of $N$ entities, each with $J$ dimensions;
            \While{stopping condition are not met}
            \For{each particle $i = 1, \dots, N$}
            \State // Set the personal best position;
            \If{$f(x_{i}) < f(y_{i})$ }
            \State $y_{i} = x_{i}$;
            \EndIf

@@

The purpose of the \index{mutation}mutation operator is to produce a trial vector for each entity in the current population by mutating a \textit{target} vector with a \textit{weighted differential}~\cite{ref:engelbrecht:2007}. The trial vector is used in the \index{crossover}crossover operator to produce offspring. The \index{mutation}mutation process then follows as such. For each parent $\boldsymbol{x}_{i}(t)$ generate a trial vector $\boldsymbol{u}_{i}(t)$ as follows:

@@

\begin{algorithm}[htb]
      \caption{The pseudo-code algorithm for the binomial \index{crossover}crossover technique for \acs{DE}.}
      \label{algo:heuristics:de:bin}
      \begin{algorithmic}
            \State $j^{*} \sim U(1,N)$;
            \State $\mathcal{J} \gets \mathcal{J} \cup \{j^{*}\}$;
            \For{each $j \in \{1, \dots, N\}$}
            \If{$U(0,1) < p_{r}$ and $j \neq j^{*}$ }
            \State $\mathcal{J} \gets \mathcal{J} \cup \{j\}$;
            \EndIf
            \EndFor
      \end{algorithmic}
\end{algorithm}

\begin{algorithm}[htb]
      \caption{The pseudo-code algorithm for the exponential \index{crossover}crossover technique for \acs{DE}.}
      \label{algo:heuristics:de:exp}
      \begin{algorithmic}
            \State $\mathcal{J} \gets \{\}$;
            \State $j \sim U(0,N - 1)$;
            \Repeat
            \State $\mathcal{J} \gets \mathcal{J} \cup \{j + 1 \}$;
            \State $j = (j+1) \mod N$
            \Until{ $U(0,1) \geq p_{r}$ or $|\mathcal{J}| = N$;}
      \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[htbp]
      \caption{The pseudo-code for the general \acs{DE} heuristic.}
      \label{algo:heuristics:de:general_de}
      \begin{algorithmic}
            \State Set the generation counter, $t = 0$;
            \State Initialise the control parameters, $\beta$ and $p_{r}$
            \While{stopping condition not met}
            \For{each entity $\boldsymbol{x}_{i}(t) \in \mathcal{C}(t)$ }
            \State Evaluate the fitness, $f(\boldsymbol{x}_{i}(t))$;
            \State Create the trial vector, $\boldsymbol{u}_{i}(t)$ by applying the mutation operator;
            \State Create an offspring, $\boldsymbol{x}'_{i}(t)$ by applying the crossover operator;
            \If{$f(\boldsymbol{x}'_{i}(t))$ is better than $f(\boldsymbol{x}_{i}(t))$ }
            \State Add $\boldsymbol{x}'_{i}(t)$ to $\mathcal{C}(t+1)$;
            \Else
            \State Add $\boldsymbol{x}_{i}(t)$ to $\mathcal{C}(t+1)$;
            \EndIf
            \EndFor
            \EndWhile
            \State Return the individual with the best fitness as the solution;
      \end{algorithmic}
\end{algorithm}

@@

Many variants of \acs{DE} have been created and researched~\cite{ref:mezura:2006}. A general notation for \acs{DE} heuristic variants have been developed by \citeauthor{ref:storn:1996}~\cite{ref:storn:1996, ref:storn:1997}. The notation follows the form \textit{DE/$x$/$y$/$z$} where $x$, $y$ and $z$ refer to the components that are used by the particular \acs{DE}. A breakdown of this notation is provided as follows:

@@

\Acp{GA} where first proposed by \citeauthor{ref:fraser:1957}~\cite{ref:fraser:1957} and later by \citeauthor{ref:bremermann:1962}~\cite{ref:bremermann:1962} and \citeauthor{ref:reed:1967}~\cite{ref:reed:1967}. However, Holland~\cite{ref:holland:1992} is widely regarded as the father of \acp{GA}. Similar to \acp{DE}, \acp{GA} are also nature-inspired population-based \acp{MH} and model genetic evolution of entities in a hypothetical population. As with other \acp{EA}, \acp{GA} implement a number of operators that drive the optimisation process. Primarily, \acp{GA} implement selection, modelling survival of the fittest and \index{crossover}crossover, modelling reproduction. The remainder of this section presents the aforementioned operators in more detail.
